Early warning signals: The charted and uncharted territories

### 2013-05-16 manuscript.tex
build tex and pdf versions

Many natural systems exhibit regime shifts - rapid changes in the state
and conditions of system behavior. Such shifts include lake
eutrophication (Carpenter, Ludwig, and Brock 1999), algal overgrowth of
coral systems (Mumby, Hastings, and Edwards 2007), fishery collapse
(Jackson et al. 2001), desertification of grasslands (Kéfi et al. 2007),
and rapid changes in climate (Dakos et al. 2008). Such dramatic shifts
have the potential to impact ecosystem health and human well-being.
Thus, it is important to develop strategies for adaptation, mitigation,
and avoidance of such shifts.

The idea that complex systems such as ecosystems could change suddenly
and without warning goes back to the 1970s (Holling 1973; May 1977).
Such early work revealed that even simple models with the appropriate
nonlinearities were capable of unpredictable behavior. The only way to
predict the transition was to have the right model -- and that meant
having already had the chance to observe the transition (even then, this
remains a tough problem). One cogent early example (Ludwig, Jones, and
Holling 1978) demonstrated how knowledge of the forms and time scales of
interactions among insects, birds, and trees could lead to a qualitative
model that essentially predicted the possibility of regime shifts.

@@

Here we review a variety of mechanisms that may lead to rapid (or
``catastrophic'') regime shifts in ecological systems, as well as
mechanisms that generate early warning signals. We focus on CSD and its
manifestations as they the most commonly studied warning signals. We
illustrate that not all rapid shifts exhibit CSD, and not all
observations of CSD involve rapid shifts. Thus the issue of determining
EWS is really two-fold: first, identify classes of systems where the
warning signal is expected and conversely systems that may undergo
shifts without such signals, and then second to determine appropriate
statistical tools to detect the warning signal. In this paper we review
both aspects of the overall question.

@@

CSD has been studied extensively in theoretical (Wissel 1984; Gandhi,
Levin, and Orszag 1998; Carpenter, Bennett, and Peterson 2006; Hastings
and Wysham 2010; Lade and Gross 2012; V. Dakos, Kéfi, et al. 2011; C.
Boettiger and Hastings 2012) and empirical contexts (Drake and Griffen
2010; J. Carpenter 2011; Veraart et al. 2011; Dai et al. 2012; Wang et
al. 2012) as a potential EWS for regime shifts. CSD occurs as a system's
dominant eigenvalue approaches zero due to a changing (possibly
deteriorating) environment. As the eigenvalue approaches zero, the
system's response to small perturbations slows. This change in dynamic
properties of a system can be expressed in greater variance,
autocorrelation, and return time of observed state variables.

@@

Much of the (most visible) recent research in EWS has focused on the
center of the diagram, where all three concepts intersect. The warning
signal patterns postulated, such as increasing variance and coefficient
of variation, (Carpenter, Bennett, and Peterson 2006), increasing
autocorrelation (Dakos et al. 2008), increasing skewness (Guttal and
Jayaprakash 2008) can all be directly derived from the changing
eigenvalue in a saddle node bifurcation. Consequently, experimental
evaluations of warning signals have largely focused on this situation as
well. CSD has frequently been studied in the context of models
exhibiting a saddle-node bifurcation.

Dai et al. (2012) maps out the entire saddle-node bifurcation diagram
empirically before presenting the corresponding decrease in recovery
time (increase in variance and autocorrelation over time). Veraart et
al. (2011) shows a very similar result in a similar system. These
important experiments are among the best demonstrations that saddle-node
bifurcation dynamics really occur in natural systems, and can be
accompanied by reliable detection of EWS, at least when sufficient data
sampling, replicates, and controls are available.

Carpenter et al. (2011) provides an even more ambitious example, in
which a lake ecosystem is manipulated towards a sudden transition
through the introduction of a predator, while a neighboring experimental
lake provides a control. In this example, the underlying dynamics of a
whole lake ecosystem are far less tractable than the laboratory
controlled chemoststats of microorganisms. While it is more difficult to
identify the bifurcation process, the system is understood well enough
to anticipate that a sudden transition can be induced under the intended
manipulation. Like the laboratory examples, this helps eliminate the
options outside the circle ``bifurcations,'' in Figure 1. The observed
warning signals then place it in the center of the diagram.

While these studies have provided valuable demonstrations of the
potential to find early warning signals of sudden transitions, this
literature has begun to enumerate examples of similar transitions in
which no such signal is present.

@@

These are some of the most problematic cases. They represent disruptive
but potentially avoidable events, but would not be detected by using CSD
as an EWS. These cases include bifurcations in continuous time
{[}@Schrieber2008{]} and discrete time {[}@Schrieber2003{]}, explicitly
spatial (Bel, Hagberg, and Meron 2012) and non-spatial, chaotic
(Schreiber 2003, ) and non-chaotic (Hastings and Wysham 2010, , )
examples. Before warning signals can be reliably applied to novel
systems, research must provide a way to discern if the dynamics
correspond to the more well understood warning signals of the
saddle-node case or the more complex patterns such as the examples
discussed here.

@@

Another example of a chaotic crisis can be found in a simple
discrete-time model where a population is subject to strong density
dependence (an Allee effect) and harvested by predators with a Type II
(saturating) functional response (Schreiber 2003). When prey have high
growth rates, the system has chaotic dynamics. Small increases in the
predation intensity cause a bifurcation with chaotic but persistent prey
populations to prey extinction. As predation intensity increases towards
this threshold, the population exhibits \emph{decreasing} variance.

\begin{figure}[htbp]
\centering
\includegraphics{schreiber-fig.pdf}
\caption{A system where variance decreases prior to a population
collapse; adapted from Schreiber (2003). In this model, prey species
with high growth rates exhibit chaotic dynamics under predation, but
populations collapse when predation increases beyond a threshold value.
Left: The population level as a function of predation rate. Middle:
Variance of the prey population level. Note that it \emph{decreases} as
predation rate approaches the threshold. Right: Lag-1 Autocorrelation in
prey population dynamics increases as the threshold is approached}
\end{figure}

Examples are not restricted to chaotic dynamics. Some spatially extended
systems exhibit a type of bifurcation that is not accompanied by CSD. In
one class of models, individual locations are subject to saddle
node-type regime shifts and influence adjacent locations via short-range
facilitation and long-range competition. Such models are used represent
transitions between vegetation types in response to changing water
availability, and reproduce naturally occurring vegetation patterns
(Rietkerk and van de Koppel 2008). In such systems, a regime shift in
one location can propagate spatially and transition the whole system
from one regime to another. Such a transition occurs if the control
parameter (e.g., rainfall), exceeds the \emph{Maxwell point} - the value
at which a local disturbance propagates outwards (Bel, Hagberg, and
Meron 2012). The Maxwell point may be far from the level at which an
individual location would undergo a saddle-node bifurcation, and thus
the system's global dynamics would not exhibit CSD prior to such a
transition. Another example is found in Schreiber and Rudolf (2008)., in
which variance is observed to decrease before a sudden transition that
follow a Hopf bifurcation .

@@

CSD precedes several types of these non-catastrophic bifurcations. In
the subcritical form of a Hopf bifurcation, a system transitions from a
stable equilibrium to a stable cycle. This bifurcation occurs in
classical predator-prey systems (May 1972) and has been observed in
experimental systems (Drake and Griffen 2010; J. Carpenter 2011; Veraart
et al. 2011; Dai et al. 2012; Wang et al. 2012). As a control parameter
approaches the critical threshold, the system's dominant eigenvalue
approaches zero and thus exhibits CSD (Chisholm and Filotas 2009; Kéfi
et al. 2012). However, the mean value of the equilibrium does not change
dramatically, and the transition from stable equilibrium to cycles is
gradual as the cycle sizes grow from zero at the threshold value. In the
presence of stochasticity, system behavior may be observed on either
side of the threshold may be indistinguishable.

@@

An experimental example a transcritical bifurcation found in Drake and
Griffen (2010), where a population of \emph{Daphnia} was forced through
a transcritical bifurcation by reducing food supplies and driving
population growth rates below zero. Indicators of CSD (variation,
skewness, autocorrelation, and spatial correlation) increased prior to
collapse of the population.

@@

\begin{figure}[htbp]
\centering
\includegraphics{kefi-fig.pdf}
\caption{A system where critical slowing down is observed without a
critical threshold, from Kéfi et al. (2012). In this model, prey have
logistic growth and are subject to predation with a Type III functional
response, but there is no bifurcation. Instead, average prey population
exhibits a smooth response to increased predation. Left: The population
level as a function of predation rate. Middle: Variance of the prey
population level. Note that it increases during the transition, despite
no bifurcation. Right: Lag-1 Autocorrelation in prey population dynamics
as predation rate increases. Note that it, too, increases despite the
lack of a bifurcation}
\end{figure}

\subsection{Non-Catastrophic Bifurcations without CSD (V)}

Some bifurcations involve neither catastrophic shifts nor patterns of
critical slowing down. The transition from very small populations to
zero is one such transition. System behavior after extinction is clearly
different, and exhibits hysteresis. However, the transition, while not
smooth, is not due to a saddle-node or similar bifurcation, and CSD
would not be expected.

\subsection{Catastrophic Regime Shifts without Bifurcations or CSD (VI)}

@@

Large, rapid changes in external conditions will result in rapid changes
in ecological system dynamics. For instance, rapid changes in North
American vegetation at the start of the Bølling-Allerød and end of the
Younger Dryas period are thought to be responses to similarly large,
rapid changes in climate (Williams, Blois, and Shuman 2011). Doney and
Sailley (2013) interpret a recent analysis by Di Lorenzo and Ohman
(2013) as demonstrating that what were previously thought of as regime
shifts in krill dynamics in the Pacific ocean (Hare and Mantua 2000)
could actually be explained by a close coupling to El Nino environmental
dynamics through the Pacific Decadal Oscillation (PDO).

Stochastic perturbations may shift systems from one regime to another
even if underlying environmental conditions remain the same. Hastings
and Wysham (2010) showed that in a model where one species with
stochastic Ricker dynamics disperses among eight patches, model behavior
can switch stochastically between wildly oscillatory behavior and
regularly cycling regimes even while parameters (including stochastic
variability) remain the same.

Schooler et al. (2011) found that lakes with the invasive plant
\emph{Salvaniai molesta} and herbivorous weevils alternated between low-
and high-\emph{Salvnia} states driven by external disturbances from
regular flooding events. Ditlevsen and Johnsen (2010) examined 25 abrupt
climate changes that occurred during the last glacial period
(Dansgaard-Oeschger events) and found no evidence for CSD in
high-resolution climate data from ice cores, and concluded that the
events were driven by endogenous climate stochasticity.

@@

@@

The above cases show that behavior providing an early warning signal
before regime shifts may only be present in certain types of ecological
systems. An additional important consideration is whether these
behaviors will be \emph{detectable} in those domains where they are
expected. To be usable as an EWS, system behavior must be detectable
well enough in advance of a regime shift to serve in decision-making,
and be reliably distinguishable from other patterns.

@@

Another approach to detecting CSD has been fitting time series data to
models. Two approaches have been used for these model-based methods.
First, models may be used to calculate summary statistics related to
CSD, such as eigenvalues (Lade and Gross 2012) or diffusion terms in
jump-diffusion models (Stephen R. Carpenter 2011; Brock and Carpenter
2012). These statistics are then examined for trends in the same fashion
as the summary statistics above. Alternatively, models representing both
deteriorating and stable conditions may be fit to the data and in order
to determine which is more likely (Dakos et al. 2012, ). Carl Boettiger
and Hastings (2012) found that likelihood ratio tests were more powerful
than trend-based summary statistic tests across several real and
simulated ecological data sets. This approach is also more robust than
summary-statistic methods to spurious correlations that arise when
collapses are driven by purely stochastic events (C. Boettiger and
Hastings 2012)

Care is required in the criteria used to judge the power of warning
signal methods. The trade-off between false negatives and false
positives is a matter of not just statistical but economic efficiency.
For instance, a large number of false positives may be acceptable if
they reduce the probability of a false warning that would result in an
otherwise avoidable catastrophic regime shift, and the costs of failing
to detect such a shift exceed that of the false positives. Carl
Boettiger and Hastings (2012) suggest the use of receiver-operating
characteristic (ROC) curves to describe the performance of various EWS.
ROC curves (Figure 5) represent the false positive rate at any true
positive rate. The area under the curve (AUC) is a useful metric of
overall performance. AUC will be one if the signal is perfect and 0.5 if
the signal performs no better than random. The complete shape of the
curve provides more information on the possible trade-offs under
different sensitivities. This information, combined with a
decision-theoretic framework, has the potential to illuminate in which
cases EWS can be useful.

@@

Recognizing the potential for early warning signals of critical
transitions represents a substantial leap forward in addressing one of
the most challenging questions in ecology and ecosystem management
today. In the decades prior, the prospect that ecosystems could make
sudden transitions into an undesirable state due to gradual, slow
changes in their environment hung like a specter over both our
understanding and management of natural systems. Research that points to
the possibility of detecting these transitions holds the promise of
meeting this challenge and has attracted justifiably widespread
attention among both theoretical and empirical communities. Nonetheless,
our understanding of early warning signals is still in its infancy. Thus
far, our best understanding and empirical experience lies in transitions
that are driven by saddle-node (also called fold) bifurcations.

While saddle-node bifurcations may be common, they represent only part
of the potential mechanisms for rapid regime shift. Occupying the center
of our diagram, Figure 1, such transitions represent our best-understood
cases. Researchers have relied on existing expertise and prior research
to identify empirical systems most likely to experience critical
transitions through the saddle-node-like mechanism (e.g. J. Carpenter
2011; Dai et al. 2012), and have achieved a close match to theoretical
predictions of early warning signals. While these examples provide a
much needed proof-of-principle that these signals can be detected in the
real world, it is too early to apply the same methods to novel systems
where the saddle-node is only one of many possible mechanisms.

Thus, establishing the saddle node mechanism a necessary condition of
using CSD as a warning signal. This can be done via manipulation in
simple experimental systems (Veraart et al. 2011; Dai et al. 2012), but
this is impractical in nature. Another approach is to assume the
saddle-node mechanism applies to a limited set of systems that have
well-studied examples, such as lakes undergoing eutrophication (Scheffer
et al. 2001), or forest/savannah transitions may also (Staver,
Archibald, and Levin 2011; Hirota et al. 2011; Bel, Hagberg, and Meron
2012). Fitting simplified saddle-node models, following Carl Boettiger
and Hastings (2012) to past regime shifts in less well-understood
systems may provide evidence for the mechanism. However, care must be
taken to specify sufficient alternative models.

CSD alone cannot be used as evidence regime shift. In some cases, it
will be present when no transition is approaching. In other cases,
regime shifts occur without CSD. Though false alarms and missed events
can occur in any statistical procedure, the cases discussed here
demonstrate that these errors will also arise when the underlying
dynamics do not correspond to our assumptions. These situations fall in
the uncharted area beyond the center of Figure 1, where research has
just begun to illuminate their existence. A better theoretical and
empirical understanding of these cases will allow us to construct novel
warning signals, that may be opposite the patterns observed in the
familiar saddle-node bifurcations. Before early warning signals can be
applied in novel systems, additional information is needed in order to
determine best signal to use.

@@

For other classes of transitions, prediction may be possible but other
EWS must be explored. Flickering (Brock and Carpenter 2010; Wang et al.
2012), or rapid transitions between states prior to a more permanent
transition, is one signal that may apply across many types of systems.
It manifests in bi-modality and high variance in times series. Spatial
pattern development may be a warning signal in systems with
short-distance positive feedbacks but long-distance negative feedbacks,
such as grassland-desert transitions (van Nes and Scheffer 2005; Kéfi et
al. 2007, ). Other spatial signals may apply where systems include both
saddle nodes and positive feedbacks across space (Guttal and Jayaprakash
2008; Litzow, Urban, and Laurel 2008; Dakos et al. 2009; Bailey 2010; V.
Dakos, van Nes, et al. 2011; Bel, Hagberg, and Meron 2012). A critical
task for EWS research is to map these signals to their domains of
applicability, and create methods to establish if ecosystems fall into
these domains.

@@

@@

>>>>

Many natural systems exhibit regime shifts - rapid changes in the state
and conditions of system behavior. Such shifts include lake
eutrophication (Carpenter, Ludwig, and Brock 1999), algal overgrowth of
coral systems (Mumby, Hastings, and Edwards 2007), fishery collapse
(Jackson et al. 2001), desertification of grasslands (Kéfi et al. 2007),
and rapid changes in climate (Dakos et al. 2008; Lenton et al. 2009).
Such dramatic shifts have the potential to impact ecosystem health and
human well-being. Thus, it is important to develop strategies for
adaptation, mitigation, and avoidance of such shifts.

The idea that complex systems such as ecosystems could change suddenly
and without warning goes back to the 1960s (Lewontin 1969; Holling 1973;
May 1977). Such early work revealed that even simple models with the
appropriate nonlinearities were capable of unpredictable behavior. The
only way to predict the transition was to have the right model -- and
that meant having already had the chance to observe the transition (even
then, this remains a tough problem). One cogent early example (Ludwig,
Jones, and Holling 1978) demonstrated how knowledge of the forms and
time scales of interactions among insects, birds, and trees could lead
to a qualitative model that essentially predicted the possibility of
regime shifts.

@@

Here we review a variety of mechanisms that may lead to rapid (or
``catastrophic'') regime shifts in ecological systems, as well as
mechanisms that generate early warning signals. We focus on CSD and its
manifestations as they are the most commonly studied warning signals. We
illustrate that not all rapid shifts exhibit CSD, and not all
observations of CSD involve rapid shifts. Thus the issue of determining
EWS is really two-fold: first, identify classes of systems where the
warning signal is expected and conversely systems that may undergo
shifts without such signals, and second, to determine appropriate
statistical tools to detect the warning signal. In this paper we review
both aspects of the overall question.

\subsection{Box: Definitions}

In this paper we refer to two closely related but different phenomena:

\textbf{Early warning signals (EWS)} - A general term for dynamic
patterns in system behavior that precede regime shifts

\textbf{Critical slowing down (CSD)} - A system's slowing response to
perturbations as it's dominant eigenvalue approaches zero, often
expressed in greater variance, autocorrelation, and return time. CSD is
one possible EWS.

@@

CSD has been studied extensively in theoretical (Wissel 1984; Gandhi,
Levin, and Orszag 1998; Carpenter and Brock 2006; Hastings and Wysham
2010; Lade and Gross 2012; V. Dakos, Kéfi, et al. 2011; C. Boettiger and
Hastings 2012) and empirical contexts (Drake and Griffen 2010; J.
Carpenter 2011; Veraart et al. 2011; Dai et al. 2012; Wang et al. 2012)
as a potential EWS for regime shifts. CSD occurs as a system's dominant
eigenvalue approaches zero due to a changing (possibly deteriorating)
environment. As the eigenvalue approaches zero, the system's response to
small perturbations slows. This change in dynamic properties of a system
can be expressed in greater variance, autocorrelation, and return time
of observed state variables.

@@

Much of the (most visible) recent research in EWS has focused on the
center of the diagram, where all three concepts intersect. The warning
signal patterns postulated, such as increasing variance and coefficient
of variation, (Carpenter and Brock 2006), increasing autocorrelation
(Dakos et al. 2008), increasing skewness (Guttal and Jayaprakash 2008)
can all be directly derived from the changing eigenvalue in a saddle
node (also called fold) bifurcation. Consequently, experimental
evaluations of warning signals have largely focused on this situation as
well. CSD has frequently been studied in the context of models
exhibiting saddle-node bifurcations.

Dai et al. (2012) studied yeast cell growth in a microcosm and
demonstrated that an Allee effect created a saddle-node bifurcation in
the system. When the cell density was reduced to levels near the
bifurcation point, a decrease in recovery time (increase in variance and
autocorrelation over time) was observed. Veraart et al. (2011) studied a
system of cyanobacteria where models suggest a saddle-node bifurcation
driven by light inhibition. They also found increases in autocorrelation
and decreased recovery rates as the system approached the bifurcation.
These important experiments are among the best demonstrations that
saddle-node bifurcation dynamics really occur in natural systems, and
can be accompanied by reliable detection of EWS, at least when
sufficient data sampling, replicates, and controls are available.

Carpenter et al. (2011) provide a larger-scale example in which a lake
ecosystem is manipulated towards a sudden transition through the
introduction of a predator, while a neighboring experimental lake
provides a control. In this and similar lake systems, bifurcation is
thought to be driven in part by trophic interactions where adult fish
prey on the competitors of their juveniles (Carpenter and Kitchell 1996;
Walters2001; Carpenter et al. 2008) which leads to a saddle-node
bifurcation. While the underlying dynamics of a whole lake ecosystem are
less tractable than the laboratory controlled chemoststats of
microorganisms, the system is understood well enough to anticipate that
a sudden transition can be induced under the intended manipulation. Like
the laboratory examples, this helps eliminate the options outside the
circle ``bifurcations,'' in Figure 1. The observed warning signals then
place it in the center of the diagram.

These studies have provided valuable demonstrations of the potential to
find early warning signals of sudden transitions. However, this
literature has begun to enumerate examples of similar transitions in
which no such signal is present.

@@

These are some of the most problematic cases. They represent disruptive
but potentially avoidable events, but would not be detected by using CSD
as an EWS. These cases include bifurcations in continuous time
{[}@Schrieber2008{]} and discrete time {[}@Schrieber2003{]}, explicitly
spatial (Bel, Hagberg, and Meron 2012) and non-spatial, chaotic
(Schreiber 2003, ) and non-chaotic (Hastings and Wysham 2010, , )
examples. Before warning signals can be reliably applied to novel
systems, research must provide a way to discern if the dynamics
correspond to the better understood warning signals of the saddle-node
case or the more complex patterns such as the examples discussed here.

@@

Another example of a chaotic crisis can be found in a simple
discrete-time model where a population is subject to strong density
dependence (an Allee effect) and harvested by predators with a Type II
(saturating) functional response (Schreiber 2003). This case is
illustrated in Figure 2. When prey have high growth rates, the system
has chaotic dynamics. Small increases in the predation intensity cause a
bifurcation with chaotic but persistent prey populations to prey
extinction. As predation intensity increases towards this threshold, the
population exhibits \emph{decreasing} variance.

\begin{figure}[htbp]
\centering
\includegraphics{schreiber-fig.pdf}
\caption{A system where variance decreases prior to a population
collapse; adapted from Schreiber (2003). In this model, prey species
with high growth rates exhibit chaotic dynamics under predation, but
populations collapse when predation increases beyond a threshold value.
Left: The population level as a function of predation rate. Mean
dynamics shown as black line, realizations with varying initial
conditions shown as grey dots; see Schreiber (2003). Middle: Variance of
the prey population level. Note that it \emph{decreases} as predation
rate approaches the threshold. Right: Lag-1 Autocorrelation in prey
population dynamics increases as the threshold is approached}
\end{figure}

Examples are not restricted to chaotic dynamics. An example is found in
Schreiber and Rudolf (2008), in which variance is observed to decrease
before a sudden transition that results in the extinction of the
population.

Another non-chaotic example is found in some spatially extended systems
that exhibit a type of bifurcation not accompanied by CSD. In this class
of models, individual locations are subject to saddle node-type regime
shifts and influence adjacent locations via short-range facilitation and
long-range competition. Such models are used represent transitions
between vegetation types in response to changing water availability, and
reproduce naturally occurring vegetation patterns (Rietkerk and van de
Koppel 2008). In such systems, a regime shift in one location can
propagate spatially and transition the whole system from one regime to
another. Such a transition occurs if the control parameter (e.g.,
rainfall), exceeds the \emph{Maxwell point} - the value at which a local
disturbance propagates outwards (Bel, Hagberg, and Meron 2012). The
Maxwell point may be far from the level at which an individual location
would undergo a saddle-node bifurcation, and thus the system's global
dynamics would not exhibit CSD prior to such a transition. This case
illustrates the importance of distinguishing between \emph{local} and
\emph{global} system dynamics and identifying the appropriate scale of
observation.

Finally, Boerlijst, Oudman, and de Roos (2013) found that indicators of
CSD do not appear prior to saddle-node bifurcations when perturbations
are not in the direction of a system's dominant eigenvalue, and even
then may only appear in one variable of the system. In their example
case, increased variance and autocorrelation only occurred when noise
was applied to the juvenile population of a model with juveniles,
adults, and predators, and it did not appear when identical noise was
applied to all three. When CSD indicators did appear, they only did so
in the juvenile population variables. This represents another
under-explored area - selecting appropriate variables for early-warning
detection in multivariate systems. Even where CSD is present, it may not
be expressed in all system components.

@@

CSD precedes several types of these non-catastrophic bifurcations. In
the subcritical form of a Hopf bifurcation, a system transitions from a
stable equilibrium to a stable cycle. As a control parameter approaches
the critical threshold, the system's dominant eigenvalue approaches zero
and thus exhibits CSD (Chisholm and Filotas 2009; Kéfi et al. 2012).
However, the mean value of the equilibrium does not change dramatically,
and the transition from stable equilibrium to cycles is gradual as the
cycle sizes grow from zero at the threshold value. To appreciate how
this bifurcation is gradual rather than catastrophic, note that in the
presence of stochasticity, the system behavior observed on either side
of the threshold may be indistinguishable: on one side stochasticity
bounces the system around a stable node, while on the other it bounces
the system around a very small limit cycle in the same region of state
space. Contrast this to a critical transition in which any stochastic
fluctuation across the threshold could lead to a qualitatively different
state.

@@

An experimental example of a transcritical bifurcation is found in Drake
and Griffen (2010), where a population of \emph{Daphnia} was forced
through a transcritical bifurcation by reducing food supplies and
driving population growth rates below zero. Indicators of CSD
(variation, skewness, autocorrelation, and spatial correlation)
increased prior to collapse of the population.

@@

\begin{figure}[htbp]
\centering
\includegraphics{kefi-fig.pdf}
\caption{A system where critical slowing down is observed without a
critical threshold, from Kéfi et al. (2012). In this model, prey have
logistic growth and are subject to predation with a Type III functional
response, but there is no bifurcation. Instead, average prey population
exhibits a smooth response to increased predation (grazing). Left: The
population level as a function of predation rate. Middle: Variance of
the prey population level. Right: Lag-1 Autocorrelation in prey
population dynamics as grazing rate increases. Note that both indicators
increase despite the lack of a bifurcation.}
\end{figure}

\subsection{Catastrophic Regime Shifts without Bifurcations or CSD (V)}

@@

Large, rapid changes in external conditions will result in rapid changes
in ecological system dynamics. For instance, rapid changes in North
American vegetation at the start of the Bølling-Allerød and end of the
Younger Dryas period are thought to be responses to similarly large,
rapid changes in climate (Williams, Blois, and Shuman 2011). Doney and
Sailley (2013) interpret a recent analysis by Di Lorenzo and Ohman
(2013) as demonstrating that what were previously thought of as regime
shifts in krill dynamics in the Pacific ocean (Hare and Mantua 2000)
could actually be explained by a close coupling to the external forcing
of El Nino environmental dynamics through the Pacific Decadal
Oscillation (PDO). Schooler et al. (2011) found that lakes with the
invasive plant \emph{Salvaniai molesta} andherbivorous weevils
alternated between low- and high-\emph{Salvnia} states driven by
disturbances from regular external flooding events. These examples
highlight cases that involve critical transitions between regimes under
circumstances that do not permit the discovery of early warning signals,
as CSD is not anticipated under these mechanisms.

Internally-driven stochastic perturbations may shift systems from one
state to another even if underlying environmental conditions remain the
same. In such conditions EWS would not be expected. Hastings and Wysham
(2010) showed that in a model where one species with stochastic Ricker
dynamics disperses among eight patches, model behavior can switch
stochastically between wildly oscillatory behavior and regularly cycling
regimes even while parameters (including stochastic variability) remain
the same. Ditlevsen and Johnsen (2010) examined 25 abrupt climate
changes that occurred during the last glacial period (Dansgaard-Oeschger
events) and found no evidence for CSD in high-resolution climate data
from ice cores, and concluded that the events were driven by endogenous
climate stochasticity rather than regime shifts {[}though see
@Cimatoribus2013 for an alternative conclusion{]}.

@@

Of course, stochastically-driven regime shifts may occur in systems
where bifurcations are also possible, and it may be difficult to
distinguish between the two. Renne et al. (2013), for example suggest
that ecosystems were under near-critical stress due to climate changes
just prior to the Chicxulub meteor impact, which resulted in mass
extinction. In such a case, EWS may precede the regime shift even if it
is ultimately triggered by a stochastic event.

@@

The above cases show that behavior providing EWS before regime shifts
may only be present in certain types of ecological systems (e.g.~see the
conditions outlined in Scheffer et al. 2009). An additional important
consideration is whether these behaviors will be \emph{detectable}. To
be usable as EWS, system behavior must be detectable well enough in
advance of a regime shift to serve in decision-making, and be reliably
distinguishable from other patterns.

@@

Another approach to detecting CSD has been fitting time series data to
models. Two approaches have been used for these model-based methods.
First, models may be used to calculate summary statistics related to
CSD, such as eigenvalues (Lade and Gross 2012) or diffusion terms in
jump-diffusion models (Stephen R. Carpenter 2011; Brock and Carpenter
2012). These statistics are then examined for trends in the same fashion
as the summary statistics above. Alternatively, models representing both
deteriorating and stable conditions may be fit to the data and in order
to determine which is more likely (Dakos et al. 2012, ). Carl Boettiger
and Hastings (2012) found that likelihood ratio tests were more powerful
than trend-based summary statistic tests across several real and
simulated ecological data sets. This approach is also more robust than
summary-statistic methods to spurious correlations that arise when
collapses are driven by purely stochastic events (C. Boettiger and
Hastings 2012).

Care is required in the criteria used to judge the power of warning
signal methods. The trade-off between false negatives and false
positives is a matter of not just statistical but economic efficiency.
For instance, a large number of false positives may be acceptable if
they reduce the probability of a false warning that would result in an
otherwise avoidable catastrophic regime shift, and the costs of failing
to detect such a shift exceed that of the false positives. Carl
Boettiger and Hastings (2012) suggest the use of receiver-operating
characteristic (ROC) curves to describe the performance of various EWS.
ROC curves (Figure 5) represent the false positive rate at any true
positive rate. The area under the curve (AUC) is a useful metric of
overall performance. AUC will be one if the signal is perfect and 0.5 if
the signal performs no better than random. The complete shape of the
curve provides more information on the possible trade-offs under
different sensitivities. This information, combined with a
decision-theoretic framework, has the potential to illuminate the cases
in which EWS can be useful.

@@

Recognizing the potential for early warning signals of critical
transitions represents a substantial leap forward in addressing one of
the most challenging questions in ecology and ecosystem management
today. In the decades prior, the prospect that ecosystems could make
sudden transitions into an undesirable state due to gradual, slow
changes in their environment hung like a specter over both our
understanding and management of natural systems. Research that points to
the possibility of detecting these transitions holds the promise of
meeting this challenge and has attracted justifiably widespread
attention among both theoretical and empirical communities. Nonetheless,
our understanding of early warning signals is still in its infancy. Thus
far, our best understanding and empirical experience lies in transitions
that are driven by saddle-node bifurcations.

While saddle-node bifurcations may be common, they represent only part
of the potential mechanisms for rapid regime shift. Occupying the center
of our diagram, Figure 1, such transitions represent our best-understood
cases. Researchers have relied on existing expertise and prior research
to identify empirical systems most likely to experience critical
transitions through the saddle-node-like mechanism (e.g. J. Carpenter
2011; Dai et al. 2012), and have achieved a close match to theoretical
predictions of early warning signals. While these examples provide a
much needed proof-of-principle that these signals can be detected in the
real world, it is too early to apply the same methods to novel systems
where the saddle-node is only one of many possible mechanisms. We are
not yet able to determine if a natural system is likely to have a
saddle-node bifurcation without detailed study, despite the popularity
of saddle-node models.

Thus, establishing the saddle node mechanism is a necessary condition of
using CSD as a warning signal. This can be done via manipulation in
simple experimental systems (Veraart et al. 2011; Dai et al. 2012), but
this is impractical in most natural systems. Another approach is to
assume the saddle-node mechanism applies to a limited set of systems
that have well-studied examples, such as lakes undergoing eutrophication
(Scheffer et al. 2001), lakes with `trophic-triangle' cascade mechanisms
(Carpenter and Kitchell 1996; Walters2001; Carpenter et al. 2008),
forest/savannah transitions (Staver, Archibald, and Levin 2011; Hirota
et al. 2011), and rangeland transitions (Walker1993; Anderies, Janssen,
and Walker 2002). Fitting simplified saddle-node models to past regime
shifts (Carl Boettiger and Hastings 2012) in less well-understood
systems may provide evidence for the mechanism. However, care must be
taken to specify sufficient alternative models.

CSD alone cannot be used as evidence of regime shifts. In some cases, it
will be present when no transition is approaching. In other cases,
regime shifts occur without CSD. Though false alarms and missed events
can occur in any statistical procedure, the cases discussed here
demonstrate that these errors will also arise when the underlying
dynamics do not correspond to our assumptions. These situations fall in
the uncharted area beyond the center of Figure 1, where research has
just begun to illuminate their existence and properties. A better
theoretical and empirical understanding of these cases will allow us to
construct novel warning signals, that may be opposite the patterns
observed in the familiar saddle-node bifurcations. Before early warning
signals can be applied in novel systems, additional information is
needed in order to determine the best signal to use.

One area that requires further exploration is the effect of different
forms of stochasticity on the existence of EWS and signal detectability.
Many processes contribute to stochastic behavior in ecological systems,
and different forms of stochasticity have different effects on system
behavior far beyond greater variance (Melbourne and Hastings 2008).
Hastings and Wysham (2010) argued that most examples of detectable CSD
indicators were found in models with additive stochasticity and smooth
potentials. Boerlijst, Oudman, and de Roos (2013), however, found that
stochasticity had the same effects whether it was additive or included
in the population growth rate. Instead, they found the \emph{direction}
of stochastic perturbations relative to the system's eigenvalue
determined whether CSD indicators were detectable. The form of
stochasticity may be important in the detectability of CSD indicators
even where CSD is present, because stochastic perturbations are needed
to explore system state-space, while at the same time can reduce the
statistical power. More work such as Perretti and Munch (2012), which
examined the role of noise color in detecting CSD, will be useful.

Another area that has is understanding how the relationship between the
scale of observation and the scale of ecological processes affect the
efficacy of EWS. As shown by the Maxwell Point example in Bel, Hagberg,
and Meron (2012), EWS which detect local bifurcations may not detect
global bifurcations in system behavior. The scale of observation likely
also will affect the statistical power of EWS. Similarly, as illustrated
in Boerlijst, Oudman, and de Roos (2013), the choice of variables to
observe in multivariate systems is important, but little is known about
how to select the appropriate variable for detecting EWS.

@@

For other classes of transitions, prediction may be possible but other
EWS must be explored. Flickering (Brock and Carpenter 2010; Wang et al.
2012), or rapid transitions between states prior to a more permanent
transition, is one signal that may apply across many types of systems.
It manifests in bi-modality and high variance in times series. Spatial
pattern development may be a warning signal in systems with
short-distance positive feedbacks but long-distance negative feedbacks,
such as grassland-desert transitions {[}@Rietkerk2004{]}. Other spatial
signals may apply where systems include both saddle nodes and positive
feedbacks across space (Guttal and Jayaprakash 2008; Litzow, Urban, and
Laurel 2008; Dakos et al. 2009; Bailey 2010; V. Dakos, van Nes, et al.
2011; Bel, Hagberg, and Meron 2012). A critical task for EWS research is
to map these signals to their domains of applicability, and create
methods to establish if ecosystems fall into these domains.

@@

Anderies, John M., Marco a Janssen, and Brian H. Walker. 2002. ``Grazing
Management, Resilience, and the Dynamics of a Fire-driven Rangeland
System.'' \emph{Ecosystems} 5 (1) (jan): 23--44.
doi:10.1007/s10021-001-0053-9.
\url{http://dx.doi.org/10.1007/s10021-001-0053-9}.

@@

Boerlijst, Maarten C., Thomas Oudman, and André M. de Roos. 2013.
``Catastrophic Collapse Can Occur without Early Warning: Examples of
Silent Catastrophes in Structured Ecological Models.'' Ed. Ricard V.
Solé. \emph{PLoS ONE} 8 (4) (apr): 62033.
doi:10.1371/journal.pone.0062033.
\url{http://dx.doi.org/10.1371/journal.pone.0062033}.
